Indicators for Ecosystem Condition in Norway

### 2023-10-04 NDVI_seminat.Rmd

<!-- RUNTIME approx. 2 min -->
```{r, eval=F}
SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != "T41")

# check if there's any 0s or 1s (which beta cannot handle)
summary(SentinelNDVI.seminat.train$mean_beta) # ok
LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
# replace 1s in Landsat data with 0.9999
LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
  mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.seminat.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region != "Northern Norway", ])
model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
summary(model.seminat.cond.Sent)

@@

```{r, include=}
model.seminat.cond.Sent <- readRDS( 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.cond.Sent.rds")
summary(model.seminat.cond.Sent)
```

@@

It is evident that there are differences between regions. But since we
are not interested in the statistics for region differences, we run
separate models for every region for ease of interpretation.

@@

```{r message=FALSE, warning=FALSE}
model.seminat.cond.Sent.N <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Northern Norway" & SentinelNDVI.seminat.train$hovedtype != "T34", ])
model.seminat.cond.Sent.C <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Central Norway", ])
model.seminat.cond.Sent.W <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Western Norway", ])
model.seminat.cond.Sent.E <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Eastern Norway", ])
model.seminat.cond.Sent.S <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Southern Norway", ])
```

@@

```{r message=FALSE, warning=FALSE}
SentinelNDVI.seminat.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.seminat.cond.Sent.N)$coefficients
```

@@

```{r message=FALSE, warning=FALSE}
SentinelNDVI.seminat.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.seminat.cond.Sent.C)$coefficients
```

@@

```{r message=FALSE, warning=FALSE}
SentinelNDVI.seminat.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.seminat.cond.Sent.W)$coefficients
```

@@

```{r message=FALSE, warning=FALSE}
SentinelNDVI.seminat.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.seminat.cond.Sent.E)$coefficients
```

@@

```{r message=FALSE, warning=FALSE}
SentinelNDVI.seminat.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.seminat.cond.Sent.S)$coefficients
```

@@

Overall take home messages:

In semi-natural ecosystems NDVI increases as ecosystem condition
deteriorates, but the effect sizes vary with major ecosystem types.
Boreal heath shows the most consistent increases, and also coastal
meadows and coastal heathlands show increases in parts of the country.
Semi-natural and semi-natural-looking meadows show very tiny increases
if any at all.

@@

```{r results='hide'}
## data handling for time series analysis
# Sentinel time series checked in exploratory analysis script

# checking time series for MODIS
ModisNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_wrap(~hovedtype)
# 2022 does not stand out as in the Sentinel data, so we keep it
ModisNDVI.seminat <- ModisNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

# checking time series for Landsat
LandsatNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_wrap(~hovedtype)
# nothing worrying to see here either
LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

```{r}
# plot
allSatNDVI.seminat %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```

@@

```{r, warning=FALSE}
model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand == "God", ])
model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$tilstand == "God", ])
model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand == "God", ])

summary(model.seminat.time.GodTilst.Sent)$coefficients$cond
summary(model.seminat.time.GodTilst.Modi)$coefficients$cond
summary(model.seminat.time.GodTilst.Land)$coefficients$cond
```

@@

Given the consistently higher NDVI-values with decreasing condition in
boreal heath, We may ask the question if the temporal NDVI increase
might be stronger under reduced condition?

```{r, warning=FALSE}
model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"), ])
model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"), ])
model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"), ])
summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T31.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T31.Land)$coefficients$cond
```

The temporal NDVI increase in boreal heath has NOT been stronger with
reduced ecological condition. So a temporal NDVI approach does not seem
like a promising way to go. We check for the same pattern in T32 and
T34, just because they are the other semi-natural major ecosystem types
with a substantial number of occurrences in the the NiN data:

```{r, warning=FALSE}
model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"), ])
model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"), ])
model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"), ])
summary(model.seminat.time.tilst.T32.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T32.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T32.Land)$coefficients$cond

@@

Again, patterns are only identifiable with Sentinel and Landsat data,
but not with Modis. Semi-natural grasslands (T32) do indeed show a
stronger increase in NDVI as condition deteriorates, so here a temporal
approach may be useful under a reduced
land-management/shrub-/tree-encroachment scenario. Coastal heathlands
(T34), on the other hand, do not show stronger temporal increases with
reduced condition. Coastal heathlands are subject to a management cycle,
which includes developmental and successional changes over a course of
15-25 years followed by prescribed burning which resets that process.
So, we have to expect increases in greenness in coastal heathlands over
time also in well-managed areas, and thus temporal changes in NDVI are
largely irrelevant for assessing ecological condition.

# Overall conclusion

NDVI in semi-natural ecosystem does vary with condition and also with
ecosystem type, but the variance and thus overlap between condition
classes and ecosystem types is large. Especially in boreal heath and
semi-natural grasslands greenness and its change over time appaer to be
indicative of the ecological condition of local ecoystems, linking well
to the registered condition variables behind the lower condition scores
(shrub/tree encroachment, low land-use intensity, succession) and thus
reduced land-use as the main pressure identified for these systems.
However, the large variance and thus overlap between condition classes
and ecosystem types makes it generally difficult to define useful
scaling values for absolute NDVI values. Since ecological condition also
relates to the temporal development of NDVI values in semi-natural
grasslands, a potential way way to approach this can be to define
scaling values for time slopes as for instance has been done for the
NDVI indicator in the ecological condition assessment for Norwegian
mountains (Framstad et al. 2022 in references above, see coded example
at <https://ninanor.github.io/IBECA/ndvi-trend_fjell.html>).

>>>>

```{r}
SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != "T41")

# check if there's any 0s or 1s (which beta cannot handle)
summary(SentinelNDVI.seminat.train$mean_beta) # ok
LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
# replace 1s in Landsat data with 0.9999
LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
  mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
```

<!-- RUNTIME approx. 2 min -->
```{r, eval=F}
# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.seminat.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region != "Northern Norway", ])
model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
summary(model.seminat.cond.Sent)

@@

```{r, include=F}
model.seminat.cond.Sent <- readRDS( 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.cond.Sent.rds")
summary(model.seminat.cond.Sent)
```

@@

It is evident that there are differences between regions. 
But since we are not interested in the statistics for region differences, 
we run **separate models for every region** for ease of interpretation.

@@

<!-- RUNTIME 5 sec -->
```{r message=FALSE, warning=FALSE}
model.seminat.cond.Sent.N <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Northern Norway" & SentinelNDVI.seminat.train$hovedtype != "T34", ])
model.seminat.cond.Sent.C <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Central Norway", ])
model.seminat.cond.Sent.W <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Western Norway", ])
model.seminat.cond.Sent.E <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Eastern Norway", ])
model.seminat.cond.Sent.S <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Southern Norway", ])
```

@@

```{r north-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Northern Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

```{r}
summary(model.seminat.cond.Sent.N)$coefficients
```

@@

```{r central-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Central Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

```{r}
summary(model.seminat.cond.Sent.C)$coefficients
```

@@

```{r west-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in W-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

```{r}
summary(model.seminat.cond.Sent.W)$coefficients
```

@@

```{r east-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in E-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

```{r}
summary(model.seminat.cond.Sent.E)$coefficients
```

@@

```{r  south-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in S-Norway."}
SentinelNDVI.seminat.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

```{r}
summary(model.seminat.cond.Sent.S)$coefficients
```

@@

> Overall take home messages:
>
> In semi-natural ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types. Boreal heath shows the most consistent increases, and also coastal meadows and coastal heathlands show increases in parts of the country. Semi-natural and semi-natural-looking meadows show very tiny increases if any at all.

@@

```{r, results='hide'}
## data handling for time series analysis
# Sentinel time series checked in exploratory analysis script

# checking time series for MODIS
   #ModisNDVI.seminat %>%
   #  group_by(id, year) %>%
   #  filter(mean == max(mean, na.rm = TRUE)) %>%
   #  ggplot(aes(x = year, y = mean)) +
   #  geom_point() +
   #  facet_wrap(~hovedtype)
# 2022 does not stand out as in the Sentinel data, so we keep it
ModisNDVI.seminat <- ModisNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

# checking time series for Landsat
   #LandsatNDVI.seminat %>%
   #  group_by(id, year) %>%
   #  filter(mean == max(mean, na.rm = TRUE)) %>%
   #  ggplot(aes(x = year, y = mean)) +
   #  geom_point() +
   #  facet_wrap(~hovedtype)
# nothing worrying to see here either
LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

```{r ndvi-time-series-comparison-ndvis, fig.cap="Comparing the NDVI time series in the three data sets."}
# plot
allSatNDVI.seminat %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```

@@

<!-- RUNTIME 10 min -->
```{r, eval=F}
model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand == "God", ])
model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$tilstand == "God", ])
model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand == "God", ])
```

```{r, include=FALSE, eval=F}
saveRDS(model.seminat.time.GodTilst.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Sent.rds")
saveRDS(model.seminat.time.GodTilst.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Modi.rds")
saveRDS(model.seminat.time.GodTilst.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Land.rds")
```

```{r, include=F}
model.seminat.time.GodTilst.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Sent.rds")
model.seminat.time.GodTilst.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Modi.rds")
model.seminat.time.GodTilst.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Land.rds")
```

```{r}
summary(model.seminat.time.GodTilst.Sent)$coefficients$cond
```

```{r}
summary(model.seminat.time.GodTilst.Modi)$coefficients$cond
```

```{r}
summary(model.seminat.time.GodTilst.Land)$coefficients$cond
```

@@

Given the consistently higher NDVI-values with decreasing condition in
**boreal heath**, We may ask the question if the temporal NDVI increase
might be stronger under reduced condition?

<!-- RUNTIME 10 min -->
```{r, warning=FALSE, eval=F}
model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"), ])
model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"), ])
model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"), ])
```

```{r, include=FALSE, eval=F}
saveRDS(model.seminat.time.tilst.T31.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Sent.rds")
saveRDS(model.seminat.time.tilst.T31.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Modi.rds")
saveRDS(model.seminat.time.tilst.T31.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Land.rds")
```

```{r, include=F}
model.seminat.time.tilst.T31.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Sent.rds")
model.seminat.time.tilst.T31.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Modi.rds")
model.seminat.time.tilst.T31.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Land.rds")
```

```{r}
summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
```

```{r}
summary(model.seminat.time.tilst.T31.Modi)$coefficients$cond
```

```{r}
summary(model.seminat.time.tilst.T31.Land)$coefficients$cond
```

The temporal NDVI increase in boreal heath has NOT been stronger with
reduced ecological condition. So a temporal NDVI approach does not seem
like a promising way to go.

We could check for the same pattern in T32 and
T34, just because they are the other semi-natural major ecosystem types
with a substantial number of occurrences in the the NiN data. Here's the code for that, stored for later.

```{r, eval=FALSE}
model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"), ])
model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"), ])
model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"), ])
summary(model.seminat.time.tilst.T32.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T32.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T32.Land)$coefficients$cond

@@

But again, patterns are only identifiable with Sentinel and Landsat data,
but not with Modis. Semi-natural grasslands (T32) do indeed show a
stronger increase in NDVI as condition deteriorates, so here a temporal
approach may be useful under a reduced
land-management/shrub-/tree-encroachment scenario. Coastal heathlands
(T34), on the other hand, do not show stronger temporal increases with
reduced condition. Coastal heathlands are subject to a management cycle,
which includes developmental and successional changes over a course of
15-25 years followed by prescribed burning which resets that process.
So, we have to expect increases in greenness in coastal heathlands over
time also in well-managed areas, and thus temporal changes in NDVI are
largely irrelevant for assessing ecological condition.

## Overall conclusion

NDVI in semi-natural ecosystem does vary with condition and also with
ecosystem type, but the variance and thus overlap between condition
classes and ecosystem types is large. Especially in boreal heath and
semi-natural grasslands greenness and its change over time appear to be
indicative of the ecological condition of local ecosystems, linking well
to the registered condition variables behind the lower condition scores
(shrub/tree encroachment, low land-use intensity, succession) and thus
reduced land-use as the main pressure identified for these systems.
However, the large variance and thus overlap between condition classes
and ecosystem types makes it generally difficult to define useful
scaling values for absolute NDVI values. Since ecological condition also
relates to the temporal development of NDVI values in semi-natural
grasslands, a potential way way to approach this can be to define
scaling values for time slopes as for instance has been done for the
NDVI indicator in the ecological condition assessment for Norwegian
mountains (Framstad et al. 2022 in references above, see coded example
at <https://ninanor.github.io/IBECA/ndvi-trend_fjell.html>).
